Owl occupancy in three of El Salvador's protected areas from 2003 to 2013

### 2021-01-12 manuscript_working/ms_working_RMD.Rmd
updated as many references as I could

--- 
title: "Owl occupancy in three of El Salvador's protected areas from 2003 to 2013"
author: "Jane West and Althea Archer"
date: "1/2/2021"
output:
  word_document:
    fig_caption: yes
    reference_docx: theme.docx
  pdf_document:
    includes:
      latex_engine: pdflatex
    keep_tex: yes
csl: ecology.csl
bibliography: owl_references.bib
---

@@

Neotropical owls are less well-studied than temperate species, such as the spotted owls of North America, and we have limited understanding of their ecological requirements, population dynamics, reproductive behavior, or conservation status [@Enriquez:2006; @Clark:1978; @Perez-Leon:2017]. Yet, neotropical owls are reliant upon ecosystems under tremendous anthropogenic pressure and rapid and unpredictable responses to climate change [@Corlett:2012]. In the tropics, particularly in rain and cloud forests, owl studies are more difficult than in temperate regions, although there are more species living in the same habitat [@Konig:1999]. In general, while our knowledge about the conservation status of neotropical populations is limited, we do know that owl abundance and distribution is decreasing in several regions, where species have been added to endangered species lists or have become locally extirpated [@Enriquez:2006].

@@

The objective of our study was to monitor population dynamics of El Salvador owls through time to determine both individual species' population dynamics and species community composition. El Salvador is located in the heart of the Mesoamerican Biodiversity Hotspot (Conservation International 2009 in Komar and Ibarra-Portillo 2009).   We surveyed owls on foot using passive listening and broadcast vocalizations in two transects each within the three largest protected areas of El Salvador from 2003 through 2013.

@@

We conducted foot surveys following the Guidelines for Nocturnaal Owl Monitoring [@Takats:2001] and a similar protocol developed for Costa Rica [@Enriquez:2001]. This study's survey protocol was an extension of previous research completed in the western half of El Imposible NP [@West:1988].  For our surveys, we created six 2-km long survey routes (transects) that each included 10 permanent survey stations at 200 m intervals. Two routes were established in each protected area (Table 1xxx).

@@

To model species richness at each route in each year, we assumed that there were 14 possible species present in El Salvador. This included the 9 species observed during our surveys and an additional 5 species that we never observed (Table 1XX). This upper limit to species richness relates directly to the 13 species known to inhabit El Salvador (cite Owls of El Salvador here) plus one species we identified (Stygian owl), which was previously undocumented in El Salvador. We thus augmented our owl detection data with 5 additional potential species with all-zero detection records [@Royle:2007]. For each species $s$ (where $s = 1,2, \ldots,15$), $w_{h,s}$ was a binary indicator of whether that owl species was present in each route based on the probability of species presence in that route, $\Omega_h$ [@Broms:2016]:

@@

@@

>>>>

--- 
title: "Owl occupancy in three of El Salvador's protected areas from 2003 to 2013"
author: "Jane West and Althea Archer"
date: "1/11/2021"
output:
  word_document:
    fig_caption: yes
    reference_docx: theme.docx
  pdf_document:
    includes:
      latex_engine: pdflatex
    keep_tex: yes
csl: ecology.csl
bibliography: owl_references.bib
---

@@

Neotropical owls are less well-studied than temperate species, such as the spotted owls of North America, and we have limited understanding of their ecological requirements, population dynamics, reproductive behavior, or conservation status [@Enriquez:2006; @Clark:1978; @Perez-Leon:2017; @Rangel-Salazar:2017]. Yet, neotropical owls are reliant upon ecosystems under tremendous anthropogenic pressure and rapid and unpredictable responses to climate change [@Corlett:2012]. In the tropics, particularly in rain and cloud forests, owl studies are more difficult than in temperate regions, although there are more species living in the same habitat [@Konig:1999]. In general, while our knowledge about the conservation status of neotropical populations is limited, we do know that owl abundance and distribution is decreasing in several regions, where species have been added to endangered species lists or have become locally extirpated [@Enriquez:2006].

@@

The objective of our study was to monitor population dynamics of El Salvador owls through time to determine both individual species' population dynamics and species community composition. El Salvador is located in the heart of the Mesoamerican Biodiversity Hotspot [@Myers:2000].   We surveyed owls on foot using passive listening and broadcast vocalizations in two transects each within the three largest protected areas of El Salvador from 2003 through 2013.

@@

We conducted foot surveys following the Guidelines for Nocturnaal Owl Monitoring [@Takats:2001], a similar protocol developed for Costa Rica [@Enriquez:2001], and raptor research guidelines [@Fuller:1987]. This study's survey protocol was an extension of previous research completed in the western half of El Imposible NP [@West:1988].  For our surveys, we created six 2-km long survey routes (transects) that each included 10 permanent survey stations at 200 m intervals. Two routes were established in each protected area (Table 1xxx).

@@

To model species richness at each route in each year, we assumed that there were 14 possible species present in El Salvador. This included the 9 species observed during our surveys and an additional 5 species that we never observed (Table 1XX). This upper limit to expected species richness relates directly to the 13 species known to inhabit El Salvador [@Dickey:1938] plus one species we identified (c.f. Stygian owl), which was previously undocumented in El Salvador. We thus augmented our owl detection data with 5 additional potential species with all-zero detection records [@Royle:2007]. For each species $s$ (where $s = 1,2, \ldots,15$), $w_{h,s}$ was a binary indicator of whether that owl species was present in each route based on the probability of species presence in that route, $\Omega_h$ [@Broms:2016]:

@@

An estimated 75% of the nearly 250 surviving species of owls are associated with dense and undisturbed forests [@Mikkola:2012].  @Perez-Leon:2017 stated that within El Salvador, natural ecosystems were the most diverse, specifically natural habitats of cloud, deciduous, riparian and pine-oak forests.  In El Salvador, 13.6% of its land is forested [@CIA:2020].  The country has continued with the traditional shade-growing coffee farming, which helps maintain forest cover.  Coffee plantations account for around 7% of its forested land because of the intensive use of shade-cover farming practices [@Silva:2016].  @Silva:2016 also stated that almost all primary or closed forest is surrounded by coffee farms, which act as a buffer zone against land use change.  @Perez-Leon:2017 stated that coffee plantations are the most important agroecosystem for owl communities’ survival in El Salvador.

@@

Distinguishing individuals by non-invasive means, such as vocal traits, is preferable when species are rare, sensitive to handling, elusive, or when other techniques are unfeasible [@Terry:2005]. Overall, the probability of detection was fairly low for all species (median values $\le 0.50$, even though we used locally recorded owl vocalizations when possible. In Guatemala, @Gerhardt:1989 found that a series of vocalizations obtained from a non-local ornithological laboratory, which were identified as vocalizations of the Mottled Owl, could not generate a response from Guatemalan owls. We found that the probability of detecting most owls increased after we broadcasted the vocalizations of Mottled Owls, Pacific Screech-owls, Crested Owls, and Spectacled Owls. On the other hand, owls were less likely to call or be detected after we played Great Horned Owl calls. Because owls, like many birds of prey, are sensitive to interactions with other owls [@Enriquez:1997; @Baumgardt:2019], we recommend that surveyors carefully consider the intra- and inter-species interactions that may affect detectability after playing specific bird vocalizations.

We did not use spectogram analysis to determine any of the specific individuals of owls that we detected. Spectograms of acoustic signals, such as owl vocalizations, are capable of not only recognizing differences between species, but even potentially individual owls of the same species or owl gender [@Eakle:1989; @Farquahar:1993; @Galeotii:1993; @Rognan:2009]. *xx can't find Eakle 1989, Farquahar 1993, Galeotii 1993* Identifying individuals within a population can improve census estimates and provide useful information on demographics, life history traits, and behavior--all of which often influence management decisions [@Terry:2005]. Spectogram analysis could help future owl researchers in El Salvadar determine more accurate population sizes of the more rare species we detected, such as Whiskered Screech-owls, c.f. Stygian Owls, and Great Horned Owls.
